---
toc: true
toc-depth: 3
echo: true
eval: false
output: false
---

# 1: R Basics

## Reading

From **R Coding Basics: An Introduction to the Basics of Coding in R** by Dr. Gaston Sanchez:

-   [First Contact with R](https://www.gastonsanchez.com/R-ice-breaker/1-04-first-contact.html)

-   [Getting Help](https://www.gastonsanchez.com/R-ice-breaker/1-05-help-documentation.html)

## Topics

-   Mathematical operations

-   Operation precedence

-   Comments

-   Objects, variables, and assignments

-   Variable names and reserved words

-   Functions

-   Help Documentation

-   Packages

## Mathematical Operations

-   Common mathematical operations in R include:

```{r}

2 + 2     # Add

3 - 4     # Subtract

5 * 2     # Multiply

4 / 2     # Divide

2 ^ 3     # Exponent
2 ** 3    # Exponent as well
```

## Operation Precedence

-   In R, operator precedence (from highest to lowest) is:

```{r}

# Parenthesis
(2 + 3) * 4

# Exponent
2 ^ 3
2 ** 3

# Multiply and divide (from left to right)
5 * 2
4 / 2

# Add and subtract
2 + 2
3 - 4
```

💻 **Hands-On**

Evaluate the following expressions in R:

-   $19 - 3 \times 6$

-   $20 - 2 \times 3^2 + 11$

-   $3 + 8 \div 2 \times 2$

-   $7 - (5 \times 3 + 2^3)$


::: {.callout-tip collapse="true"}
## Answer

```{r}
#| eval: true
#| output: true

19 - 3 * 6

20 - 2 * 3^2 + 11

3 + 8 / 2 * 2

7 - (5 * 3 + 2^3)
```
:::

## Comments

-   Comments are notes to explain what the code does.

-   In R, comments are created using a hash `#`.

-   On a line, everything after a hash will be ignored. In other words, even if the comments contain some code, it will not be executed.

-   The keyboard short cut to comment out blocks of code is `Command` or `Ctrl` + `Shift` + `c`

```{r}

(2 + 3) * 4    # 2 + 3 will be evaluated first, then multiplied by 4
```

💻 **Hands-On**

Run the following code in the console and add a comment describing the output.

```{r}

5 * (2 + 4) / 2 

(5 * 2 + 4) / 2

5 * (2 + 4 / 2)
```

::: {.callout-tip collapse="true"}
## Answer

```{r}
#| eval: true
#| output: true

5 * (2 + 4) / 2    # 15

(5 * 2 + 4) / 2    # 7

5 * (2 + 4 / 2)    # 20
```
:::

## Object, Variable, and Assignment

-   An **object** is anything that stores values or instructions.

-   A **variable** is a name associated to an object so that we can access the object in memory.

-   A variable can be created using the assignment operator `<-` (left arrow) or the equal sign `=`.

```{r}

# 100 is an object itself
100

# We name this object 'eagles'
eagles <- 100

# The variable 'eagles' allows us to access the object 100
eagles
```

💻 **Hands-On**

Create the following variables:

-   `year` that stores the value `2026`

-   `price` that stores the value `5.5`

-   `score` that stores the value `93`

::: {.callout-tip collapse="true"}
## Answer

```{r}
#| eval: true
#| output: true

year <- 2026

price <- 5.5

score <- 9.3
```
:::

## Variable Names

-   Variable names are **case-sensitive**.

```{r}

# These names are NOT the same
eagles

Eagles

EAGLES

eaGleS

EagLEs

EaGlEs
```

💻 **Hands-On**

Run the following code in the console. What does it return?

```{r}

EAGLES <- 123

Eagles
```

::: {.callout-tip collapse="true"}
## Answer

Since variable names are case-sensitive, `EAGLES` and `Eagles` are different. In this case, we created `EAGLES` but not `Eagles` so R will throw an error.

```{r}
#| eval: true
#| output: true
#| error: true

EAGLES <- 123

Eagles
```
:::

## Variable Names

-   A variable name must start with a letter or a period (not recommended). The rest can only consist of letters `a-z, A-Z`, numbers `0-9`, underscores `_`, and periods `.`

-   A variable name cannot be a reserved word (`TRUE`, `FALSE`, `if`, `for`, `while`, etc.)

```{r}

# Valid variable names
age            
height_cm1    
weight2.kg    

# Valid but not recommended
.data       
True
```

## Reserved Words

-   **Reserved words** are words with special meaning and [cannot]{.underline} be used as variable names.

```{r}

if, else, for, while, function 

break, stop, continue, next

TRUE, FALSE

NULL, NA, Inf, NaN
```

💻 **Hands-On**

The following R code attempts to create variables. Run each line in the console and explain why the variable name is invalid.

```{r}

9null <- 5

null%9 <- 5

null 9 <- 5

_null9 <- 5 

null,9 <- 5

NULL <- 5
```

::: {.callout-tip collapse="true"}
## Answer

Note that all given variable names are invalid. See below for the explanation.

```{r}

9null <- 5     # variable name cannot start with a number

null%9 <- 5    # variable name cannot contain any percentage

null 9 <- 5    # variable name cannot contain any space

_null9 <- 5    # variable name cannot start with an underscore

null,9 <- 5    # variable name cannot contain any comma

NULL <- 5      # variable name cannot be a reserved word
```
:::

## Naming Styles

Three common naming styles for variables include

```{r}

eagles_green_swag    # snake case

eaglesGreenSwag      # camel case

eagles.green.swag    # period case
```

💻 **Hands-On**

Create three variables that represent the phrase "Crimson Roll Tide" using the three naming styles.

::: {.callout-tip collapse="true"}
## Answer

[Tidyverse style guide]{https://style.tidyverse.org/syntax.html) recommends the snake case, which is what I tend to use as well!

```{r}

crimson_roll_tide    # snake case

crimsonRollTide      # camel case

crimson.roll.tide    # period case
```
:::

## **Functions**

-   A function is a collection of R code to perform a complex task or obtain certain results.

-   R has many useful built-in functions.

```{r}

# Absolute value, sign function, and squared root
abs, sign, sqrt

# Rounding of numbers
round, floor, ceiling, trunc, signif

# Logarithms and Exponentials
exp, log, expm1, log1p

# Combination and factorial
choose, factorial, lchoose, lfactorial

# Trigonometric and hyperbolic functions
cos, sin, tan, cospi, sinpi, tanpi, acos, asin, atan
cosh, sinh, tanh, acosh, asinh, atanh

# Beta and gamma functions
beta, lbeta, gamma, lgamma, digamma, trigamma, psigamma
```

💻 **Hands-On**

Try the following code in the console.

```{r}

abs(-7)         # absolute value

sqrt(25)        # squared root

exp(3)          # exponential

log(8)          # natural log

factorial(6)    # 1 x 2 x 3 x 4 x 5 x 6
```

::: {.callout-tip collapse="true"}
## Answer

```{r}
#| eval: true
#| output: true

abs(-7)         # absolute value

sqrt(25)        # squared root

exp(3)          # exponential

log(8)          # natural log

factorial(6)    # 1 x 2 x 3 x 4 x 5 x 6
```
:::

## Getting Help

-   If we want to learn more about a function, we can refer to its **help documentation**.

```{r}

# Using the operator '?'
?abs
?sqrt

# Using the help() function
help(abs)
help(sqrt)
```

💻 **Hands-On**

Use the help documentation to explain the purpose of the following functions.

```{r}

sign(-17)

log1p(3)

choose(5, 2)
```

::: {.callout-tip collapse="true"}
## Answer

To get the help documentation of the functions, we can use `?` or `help()`

```{r}
?sign
help(sign)

?log1p
help(log1p)

?choose
help(choose)
```

The functions do the following

```{r}
#| eval: true
#| output: true

sign(-17)       # give -1 for negative number and 1 for positive number

log1p(3)        # compute log(1 + 3)

choose(5, 2)    # compute combination coefficient 5C2 
```
:::

💻 **Hands-On**

What does the following code do? Use the help documentation of `round()` to confirm your guess/observation.

```{r}

round(6.267)

round(6.267, digits = 1)
round(6.267, 1)

round(6.267, digits = 2)
round(6.267, 2)
```

::: {.callout-tip collapse="true"}
## Answer

To get the help documentation of `round()`, we can use `?` or `help()`. It is actually pretty self-explanatory! 

```{r}
?round
help(round)
```

The given code does the following

```{r}
#| eval: true
#| output: true

# Round to the whole number (digits = 0)
round(6.267)

# Round to 1 decimal place
round(6.267, digits = 1)

# Round to 2 decimal places
round(6.267, digits = 2)
```
:::

## Packages

-   A **package** is a collection of functions and data for specific purpose.

-   Most developers make their packages freely available on the Comprehensive R Archive Network (CRAN).

```{r}

# Install a package
install.packages("bannerCommenter")

# Load an installed package
library(bannerCommenter)
library("bannerCommenter")

# Uninstall a package
remove.packages("bannerCommenter")
```

<!-- ## Exercise 1 -->

<!-- Evaluate the expression at $x = 3$, $y = -2$, and $z = 4$ -->

<!-- -   $6z + 5x - 3y$ -->

<!-- -   $x^2 - 5y + 4z$ -->

<!-- -   $\dfrac{4z - 2y}{20 x}$ -->

<!-- ::: {.callout-tip collapse="true"} -->
<!-- ## Answer -->

<!-- ```{r} -->


<!-- ``` -->
<!-- ::: -->

<!-- ## Exercise 2 -->

<!-- The balance $A$ in an account that earns an annual compound interest rate $r$ (in decimal form) for $t$ years is given by -->

<!-- $$ -->
<!-- A = P \left( 1 + \dfrac{r}{n} \right)^{nt}, -->
<!-- $$ -->

<!-- where $P$ is the principal (initial deposit) and $n$ is the number of compoundings per year. -->

<!-- 1.  We deposit $\$6000$ in an account with an annual interest rate of $1.5\%$. What is our balance after $5$ years when the interest is compounded quarterly ($n = 4$)? -->

<!-- 2.  We deposit $\$10,000$ in an account with an annual interest rate of $3.75\%$. What is our balance after $12$ years when the interest is compounded daily ($n = 365$)? -->

<!-- ::: {.callout-tip collapse="true"} -->
<!-- ## Answer -->

<!-- ```{r} -->


<!-- ``` -->
<!-- ::: -->

<!-- ## Exercise 3 -->

<!-- The compound interest formula can be rewritten as -->

<!-- $$ -->
<!-- P = \dfrac{A}{\left( 1 + \dfrac{r}{n} \right)^{nt}} -->
<!-- $$ -->

<!-- Find the principal $P$ that we need to invest today at $4.5\%$ annual interest compounded monthly to earn $\$100,000$ in the next $20$ years. -->

<!-- ::: {.callout-tip collapse="true"} -->
<!-- ## Answer -->

<!-- ```{r} -->


<!-- ``` -->
<!-- ::: -->

<!-- ## Exercise 4 -->

<!-- The formula for the volume of a cylinder is -->

<!-- $$ -->
<!-- V = \pi r^2 h -->
<!-- $$ -->

<!-- where $r$ is the radius and $h$ is the height. -->

<!-- Find the height of a cylindrical shipping container with a volume of $15,607$ cubic inches and a radius of $12$ inches. You may find reading `help(pi)` or `?pi` useful. -->

<!-- ::: {.callout-tip collapse="true"} -->
<!-- ## Answer -->

<!-- ```{r} -->


<!-- ``` -->
<!-- ::: -->
